Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming

### 2025-06-07 paper/draft.tex
Add explanation of xclingo excerpt

\maketitle
\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction system 
that leverages the strengths of both paradigms. 
This approach shows strong performance on real patient data.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

@@

Since patients may not exhibit \textit{all} known symptoms, 
the system must support reasoning under partial information. 
To allow flexible reasoning, a choice rule introduces possible symptoms 
into the knowledge base when patient data is incomplete:
\begin{equation}
    \{ add(symptom(S)) : symptom(S) \}.
\end{equation}

@@

Symbolic artificial intelligence offers a significant advantage 
over subsymbolic methods in terms of explainability. 
In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
enabling transparent reasoning. However, ASP solvers such as \textit{Clingo} do not, 
at the time of writing,\footnote{At the time of writing this paper.} 
provide intuitive traces of their reasoning processes. 
Prior work has explored methods for associating logical propositions 
with their justifications \cite{cabalar2014causal}. For example, 
the logic program shown below is accompanied by a causal graph 
in \cref{fig:causal-g}, which illustrates the connections 
between rules and their inferred outcomes. 
This structure supports the interpretability of conclusions derived by a symbolic logic solver.
\begin{equation}
\begin{aligned}[b]
    l :& punish \longleftarrow drive, drunk \\
    m :& punish \longleftarrow resist \\
    e :& prison \longleftarrow punish \\
    d :& drive \\
    k :& drunk \\
    r :& resist \\
\end{aligned}
\label{eq:example-cg}
\end{equation}

\begin{figure}
    \centering
    \includegraphics{assets/causal_g.png}
    \caption{Causal graph of logic program (\cref{eq:example-cg}).}
    \label{fig:causal-g}
\end{figure}

The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} 
allows the presentation of justifications regarding the results given by the solver \textit{Clingo}. 
This is achieved by adding a \src{trace} corresponding to the logic program's rules. 
roviding a clear explanation of how the system reached specific conclusions can help identify potential 
errors in the creation of the knowledge base and provide the system user with a rationale for the final outcome. 
This is particularly important when the application is used in medical contexts.

@@

Evaluation was conducted using publicly available patient data.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}} 
Each patient's symptoms were provided as input to the system. 
The knowledge base combines all rules derived from the encoded medical literature. 
A prediction is considered successful when the disease identified by the ASP solver matches the patient’s actual diagnosis.
\begin{table}[h]
    \centering
    \begin{tabular}{lcc}
    \hline
    Diseases & Size of Knowledge Base & Accuracy \\
    \hline
    \hline
    Chicken Pox      &  66  & 95\% \\
    Pneumonia        &  75  & 100\% \\
    Common Cold    &   44 & 100\% \\
    \hline
    \end{tabular}
    \caption{Accuracy of the method on selected diseases.}
\end{table}

The model achieves high accuracy when a sufficiently 
large set of rulesis included in the knowledge base. 
The size of the knowledge base is defined as the number 
of logical terms in the ASP program associated with each disease.

The system also incorporates an explanation mechanism using the \src{xclingo} tool,
which generates a graph visualizing the logical associations used to derive the diagnosis.

\begin{lstlisting}[caption={A part of the explanation for the chickenpox diagnosis. 
    The decision tree illustrates the symptom associations and outlines 
    the reasoning process used by the system to reach its conclusion.},
label=lst:terms]{Name}
  *
  |__diagnosis(chickenpox)
  |  |__has(symptom(itching));has(symptom(itching))
  |  |__has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue))
  |  |__has(symptom(lethargy))
  |  |  |__has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue))
  |  |  |__linked_symptom(fatigue,lethargy)
  |  |__has(symptom(high_fever))
  |  |  |__has(symptom(mild_fever));has(symptom(mild_fever))
  |  |  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
  |  |  |  |__linked_symptom(loss_of_appetite,mild_fever)
  |  |  |__linked_symptom(mild_fever,high_fever)
  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
  |  |__has(symptom(mild_fever));has(symptom(mild_fever))
  |  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
  |  |  |__linked_symptom(loss_of_appetite,mild_fever)
  |  |__has(symptom(swelled_lymph_nodes))
  |  |  
  ...
\end{lstlisting}

\section{Conclusions}

This work presents a method for automatic disease prediction 
that encodes medical literature and transforms it into ASP code using an LLM. 
It describes the roles of both LLMs and ASP, emphasizing how each contributes to the overall system. 
The system architecture is outlined, along with implementation details for constructing effective prompts 
and structuring rules within the ASP knowledge base.
The method includes predictions for a set of diseases based on the constructed knowledge bases. 
Results show that this approach achieves high-accuracy predictions and provides interpretable explanations.

>>>>

\maketitle
\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction framework 
that leverages the strengths of both paradigms. 
This approach shows strong performance on real patient data.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

@@

Since patients may not exhibit \textit{all} known symptoms, 
\sys must support reasoning under partial information. 
To allow flexible reasoning, a choice rule introduces possible symptoms 
into the knowledge base when patient data is incomplete:
\begin{equation}
    \{ add(symptom(S)) : symptom(S) \}.
\end{equation}

@@

Symbolic artificial intelligence offers a significant advantage 
over subsymbolic methods in terms of explainability. 
In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
enabling transparent reasoning. However, ASP solvers such as \textit{Clingo} do not, 
at the time of writing,\footnote{At the time of writing this paper.} 
provide intuitive traces of their reasoning processes. 
Prior work has explored methods for associating logical propositions 
with their justifications \cite{cabalar2014causal}. For example, 
the logic program shown below is accompanied by a causal graph 
in \cref{fig:causal-g}, which illustrates the connections 
between rules and their inferred outcomes. 
This structure supports the interpretability of conclusions derived by a symbolic logic solver.
\begin{equation}
\begin{aligned}[b]
    l :& punish \longleftarrow drive, drunk \\
    m :& punish \longleftarrow resist \\
    e :& prison \longleftarrow punish \\
    d :& drive \\
    k :& drunk \\
    r :& resist \\
\end{aligned}
\label{eq:example-cg}
\end{equation}
% https://docs.google.com/drawings/d/1Ing4GFH2kbniiHI5rbXzw-xcAUQAxWAJpgmg4tz1fzc/edit?usp=sharing
\begin{figure}
    \centering
    \includegraphics{assets/causal_g.pdf}
    \caption{Causal graph of logic program (\cref{eq:example-cg}).}
    \label{fig:causal-g}
\end{figure}

The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}}~\cite{Cabalar_2020} 
enables the visualization of justifications for results produced by the \textit{Clingo} solver. 
It does so by incorporating a \src{trace} that corresponds to the rules of the logic program. 
Providing clear explanations of how the framework reaches specific conclusions helps users identify 
potential errors in the construction of the knowledge base and offers transparency for the final outcome. 
This is especially important in medical contexts, where interpretability and trust are critical.

@@

Evaluation was conducted using publicly available patient data.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}
Each patient's list of symptoms serves as input.  
The knowledge base—comprising ASP rules automatically derived  
from medical literature—infers potential diagnoses.  
A prediction is considered correct when the disease inferred  
by the ASP solver matches the actual diagnosis listed in the dataset.

\begin{table}[h]
    \centering
    \begin{tabular}{lcc}
    \hline
    Diseases & Size of Knowledge Base & Accuracy \\
    \hline
    \hline
    Chicken Pox    &  66  & 95\% \\
    Pneumonia      &  75  & 100\% \\
    Common Cold    &  44  & 100\% \\
    \hline
    \end{tabular}
    \caption{Accuracy of the framework on selected diseases. 
    The size column indicates the number of logical terms in the disease-specific ASP program.}
\end{table}

High accuracy is achieved when the knowledge base includes a sufficiently 
rich set of logical rules. The size of this base, measured as the number 
of ASP terms associated with a disease, correlates with prediction performance.

In addition to accurate predictions, \sys offers explainability 
through integration with \src{xclingo}. 
This tool visualizes the reasoning process behind each diagnosis, 
revealing the logical path from symptoms to conclusion.

\begin{lstlisting}[caption={Excerpt from the explanation tree for the chickenpox diagnosis. The structure illustrates the reasoning path based on symptom associations.},
    label=lst:chickenpox_explanation]{Name}
    *
    |__ diagnosis(chickenpox)
        |__ has(symptom(itching))
        |__ has(symptom(fatigue))
        |__ has(symptom(lethargy))
            |__ has(symptom(fatigue))
            |__ linked_symptom(fatigue, lethargy)
        |__ has(symptom(high_fever))
            |__ has(symptom(mild_fever))
                |__ has(symptom(loss_of_appetite))
                |__ linked_symptom(loss_of_appetite, mild_fever)
            |__ linked_symptom(mild_fever, high_fever)
        |__ has(symptom(loss_of_appetite))
        |__ has(symptom(mild_fever))
            |__ has(symptom(loss_of_appetite))
            |__ linked_symptom(loss_of_appetite, mild_fever)
        |__ has(symptom(swelled_lymph_nodes))
\end{lstlisting}

\Cref{lst:chickenpox_explanation} shows an excerpt from the explanation 
tree generated for a patient diagnosed with \textit{chickenpox}. 
The tree visualizes the reasoning steps performed by the ASP solver to reach this conclusion. 
At the root, \sys infers \texttt{diagnosis(chickenpox)} 
based on the presence of multiple symptoms such as \texttt{itching}, \texttt{fatigue}, \texttt{lethargy}, 
and \texttt{high\_fever}. 
Each symptom is further justified through observed inputs or inferred relationships. 
For example, \texttt{lethargy} is supported by repeated instances of \texttt{fatigue} 
and the known association \texttt{linked\_symptom(fatigue, lethargy)}. 
Similarly, \texttt{high\_fever} is explained via the intermediate symptom \texttt{mild\_fever}, 
which itself is supported by \texttt{loss\_of\_appetite}. 
These explanations demonstrate how \sys combines direct observations 
and transitive symptom relationships encoded in the knowledge base to 
construct a transparent and interpretable diagnostic path.

\section{Conclusion}

This work presents a method for automatic disease prediction 
that encodes medical literature and transforms it into ASP code using an LLM. 
It describes the roles of both LLMs and ASP, emphasizing how each contributes to the overall framework. 
The \sys architecture is outlined, along with implementation details for constructing effective prompts 
and structuring rules within the ASP knowledge base.
The method includes predictions for a set of diseases based on the constructed knowledge bases. 
Results show that this approach achieves high-accuracy predictions and provides interpretable explanations.
